BIP Architectures and Open pNets

### 2019-01-31 2018-Arch-pNets/arch-pNets.tex
switch pnet and archi
Imported biblio from net paper
Notations in 1 page, we can remove more I believe

\section{Previous results: Architecture and pNets}

\subsection{Preliminaries}
\label{secn:preliminaries}

@@

\subsubsection{Notations}
\label{secn:notations}

We extensively use indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. 
Thus, 
$a_i^{i\in I}$
%, or equivalently  $(i\mapsto a_i)^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$. % Such a family
% is equivalent to the mapping $(i\mapsto a_i)^{i\in I}$.
% To specify the set over which the structure is indexed,
% indexed structures are always denoted with an exponent of the form $i\in I$
% (arithmetic only appears in the indexes if necessary).
This notation defines both $I$ the set over which the family is
indexed (called \emph{range}), and $a_i$ the elements of the family.
E.g., $a^{i\in\{3\}}$ is the mapping with a single entry $a$ at index
$3$ ; abbreviated $(3\mapsto a)$ in the following.
When this is not
ambiguous, we shall use notations for sets, and typically write
``indexed set over I'', even though formally we should speak of multisets; and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$. We use a classical overline notation, \eg $\overline{a}$, 
for a family, where the indexing set is irrelevant. Finally, we use $\uplus$ to denote 
the disjoint union on
indexed sets.

\noteIn{states looks ok, but initial state is $s_0$ in pNet and not in archi : $s^0$.  Exponent 
position was removed from pNet because we found it complex with indexed set notation but 
could be ok.
\Ludo

I am using exponent notation for ``modifiers'', reserving\mdash
as much as possible\mdash index notation for enumeration.  I
think this can be solved, where necessary, by using parentheses,
\eg $(s^0_i)^{i \in I}$
\Simon
}

\subsubsection{Term algebra}
\label{secn:terms}

\todoLHin{Remove term algebra or simplify?}
\todoLHin{universal domain $\implies$ remove sort}
\todoLHin{replace assignments by $e_i$

  In fact, in the pNet action terms, you need expressions, not
  assignements.  Therefore assuming that all expressions are
  assignments will not work, so the right approach is to have
  expressions (for pNets) and assignments (for BIP) separately.
  \Simon
%
}

\noteSBin{In Archi we have $a$, whereas in pNets we have $\alpha$, where, approximately $\alpha = 
a(x_1..x_n)$, except that $a$ is port name in archi and action name in pNets.
Is this true? can we unify better? or is it sufficient?
\Ludo

$a$ is a set of ports, but otherwise, yes, this is true.  I think
this is coherent, since, in archi, we have the ``interaction
part'', \ie only ports, plus the data part, \ie variables,
guards, expressions.  In pNets, these are combined.  So, it seems
ok that notations differ\mdash ``uniformisation'' will be
achieved by encoding archi $\rightarrow$ pNets.
}

\noteLHin{I believe we should simplify this paragraph about param actions

  I agree with this, but also we may look at what is useful here, and what is only used later in pLTS/pNet definitions. 
\Eric}
The pNet model in \secn{pNets} relies on a notion of parameterised actions, that are
symbolic expressions using data types and variables. As we aim at encoding the low-level 
behaviour of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on (some sort of) unification between two or more action
expressions, to express various kinds of communication or
synchronisation mechanisms.

%\def\Talg{\mathcal{T}_{\Sigma,\P}}  % Moved to the header

\noteSBin{proposed notations:

  $g$ are boolean expressions;
  $e_i$ are updates/assignments
  \Ludo

  In fact, I am not sure whether I have correctly understood this
  suggestion: do you mean, \eg using $g$ to denote the set of
  Boolean expressions or individual Boolean expressions? Using
  lower case letters to denote sets might be problematic, since
  we (at least I) use them a lot to denote instances, as in
  \[q \goesto{a, g, e} q'\,.\]
  %% In fact, I even use capital letters: in the definition of
  %% interaction model composition, I use
  %% \[\cdot \goesto{a, G, E} \cdot\]
  %% in the conclusion of the rule.  However, for this latter I can
  %% easily change the notation.

  I have set up macros for, general and Boolean expressions,
  actions, variables \etc[,] which I have substituted everywhere,
  where appropriate in the following paragraph.
}

  \begin{tabular}{
      @{}p{0.15\columnwidth}@{}
      p{0.35\columnwidth}@{}
      p{0.5\columnwidth}@{}
    }
    {\bf Output} & {\bf Command} & {\bf Definition}
    \\
    $\data$ &
    {\ttfamily\textbackslash data} &
    {\ttfamily\textbackslash sD}
    \\
    \signature &
    {\ttfamily\textbackslash signature} &
    {\ttfamily\textbackslash Sigma}
    \\
    \variables &
    {\ttfamily\textbackslash variables} &
    {\ttfamily\textbackslash cV}
    \\
    \Talg &
    {\ttfamily\textbackslash Talg} &
    {\ttfamily\textbackslash cT\_\{\textbackslash signature,\textbackslash variables\}}
    \\
    \actions{\#1} &
    {\ttfamily\textbackslash actions\{\#1\}} &
    {\ttfamily\textbackslash cA\_\{\#1\}}
    \\
    \exprs{\#1} &       
    {\ttfamily\textbackslash exprs\{\#1\}} &
    {\ttfamily\textbackslash sE\_\{\#1\}}
    \\
    \monexprs{\#1} &       
    {\ttfamily\textbackslash monexprs\{\#1\}} &
    {\ttfamily\textbackslash sE\^{}\{<\}\_\{\#1\}}
    \\
    \boolexprs{\#1} &       
    {\ttfamily\textbackslash boolexprs\{\#1\}} &
    {\ttfamily\textbackslash sB\_\{\#1\}}
    \\
    \guards{\#1} &
    {\ttfamily\textbackslash guards\{\#1\}} &
    {\ttfamily\textbackslash sB\^{}\{<\}\_\{\#1\}}
    \\
    \assigns{\#1} &
    {\ttfamily\textbackslash assigns\{\#1\}} &
    {\ttfamily\textbackslash sA\_\{\#1\}}
    \\
    \updates{\#1} &
    {\ttfamily\textbackslash updates\{\#1\}} &
    {\ttfamily\textbackslash sA\^{}\{<\}\_\{\#1\}}
    \\
    \valuations{\#1} &
    {\ttfamily\textbackslash valuations\{\#1\}} &
    {\ttfamily\textbackslash data\^{}\{\#1\}}
    \\[6pt]
    \val[]{\#2}{\#3} &
    {\ttfamily\textbackslash val[]\{\#2\}\{\#3\}} &
    {\ttfamily\#1\{\textbackslash sigma\}\^{}\{\#2\}\_\{\#3\}}
    \\[6pt]
    \val[\tilde]{\#2}{\#3} &
    {\ttfamily\textbackslash val[\textbackslash tilde]\{\#2\}\{\#3\}} &
    (optional argument usage examples)
    \\[6pt]
    \val[\primeit]{\#2}{\#3} &
    {\ttfamily\textbackslash val[\textbackslash primeit]\{\#2\}\{\#3\}} &
    \\[6pt]
    \val[\secondit]{\#2}{\#3} &
    {\ttfamily\textbackslash val[\textbackslash secondit]\{\#2\}\{\#3\}} &
  \end{tabular}
\noteSBin{commands above (tabular generates an error inside a
  note)}

Formally, we assume the existence of a term algebra $\Talg$,
where $\signature$ is the signature of the data and action constructors,
and $\variables$ a set of \emph{variables}. Within $\Talg$, we distinguish a set of
\emph{data expressions} $\exprs{\variables}$ and a set of \emph{Boolean
expressions} \todoLH{$\mathcal{B} \to g$}
$\boolexprs{\variables}\subseteq\exprs{\variables}$.
On top of $\exprs{\variables}$ we build the \emph{action algebra}
$\actions{\variables}$, with $\actions{\variables}\subseteq\Talg$ \replaceSB{,}{and}
$\exprs{\variables}\cap\actions{\variables} = \emptyset$;
naturally action terms will use data expressions as subterms.
\addSB{In addition to the above notations, we will use
\[
\assigns{\variables} \bydef{=}
\Setdef{(x_i := e_i)^{i \in I}}{
  x_i^{i \in I}\subseteq \variables,
  e_i^{i \in I}\subseteq \exprs{\variables},
  I \text{ is a finite index set}
}
\]
to denote the set of \emph{variable assignments}.}  We use the symbol
$\noop$ to denote the empty assignment, \ie an assignment with
$I = \emptyset$.

To be able to reason about the data flow between pLTSs, we
distinguish \emph{input variables} of the form $?x$ within terms; the function
$\vars(t)$ identifies the set of variables in a term
$t\in\Talg$, and $iv(t)$ returns its input variables.
Action algebras can naturally encode usual point-to-point message passing calculi (using 
$a(?x_1,\dots,?x_n)$ for inputs, $a(v_1,\dots,v_n)$ for outputs), but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.

Finally, we assume the existence of a universal data domain
given as a partially-ordered set $(\data, \order)$, potentially
encompassing several copies of any given data type with
different orders.  For example, we assume that $(\data,
\order)$ comprises both the unordered set of Booleans $\sB =
\tuple{\{\true, \false\}, \emptyset}$ and the naturally ordered one
$\sB^< = \tuple{\{\true, \false\}, \{\false < \true\}}$, and
similarly for integer and real numbers; as well as the set of
intervals ordered by inclusion.
%
\addSB{In practice, and in the examples of
this paper, all the variables are sorted, meaning that their values
can only range over given subsets of $\data$, such as $\sB$ or
$\sB^<$.
%
When speaking of an ordered sort, \eg $\sB^<$, we will assume that it
forms a meet-semilattice and denote by $\wedge$ the corresponding meet
operator.}

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$.
Valuations extend canonically to expressions.  Thus
$\val{}{}(x)$ and $\val{}{}(e)$ denote, respectively, the
values of the variable $x$ and the expression $e$ under the
valuation $\val{}{}$.
%
\addSB{Since we consider variables to be sorted, valuations
  applied to expressions are interpreted as partial functions
  $\exprs{V} \rightharpoonup \data$.}

For a valuation $\val{}{} \in
\valuations{V}$ and an assignment $(x_i := e_i)^{i \in I} \in
\assigns{V}$, we denote $\val{}{}[(x_i := e_i)^{i \in I}]$ the
valuation defined by
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e \in \exprs{V}$ is called \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{} \in \valuations{V}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} \in \assigns{V}$ is
called monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We will denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%

\subsection{The theory of architectures with data}
\label{secn:archs}

@@

@@

\bibliographystyle{abbrv}
\bibliography{biblio}

@@

\begin{definition}[Sorts, Holes, Leaves of pNets]
  \begin{itemize}
  \item  \noteSB{I think this can be lightened up a bit, since the definitions are all 
  straightforward, whereas the notation is complex.  Maybe reformulate in words and move 
  the formal definitions in the appendix?}
 The sort of a pNet is its signature, i.e. the set of actions it can
perform. \noteSB{Is there any chance that, by writing assignments explicitly ($x := e$), 
we could eliminate the need to distinguish the input variables at all? That would 
simplify things.}\hl{In the definition of sorts, we do not need to distinguish
input variables} (that specify the dataflow within LTSs), so for
computing LTS sorts, we use a substitution operator\footnote{$\subst{y_k\gets x_k}^{k\in 
K}$ is the parallel substitution 
operation.} to remove the
\emph{input marker} of variables. Formally:
\[
\begin{array}{l}
\Sortop(\mylangle S,s_0, \to\myrangle) = \{\alpha\subst{x \gets ?x| 
x\in\symb{iv}(\alpha)}|s \xrightarrow{\langle \alpha,~g,~(x_j\!:= {e}_j)^{j\in
    J}\rangle} s'\in \to \} \\
\Sortop(\mylangle \overline{\pNet}\!, %\pNet_i^{i\in I}, \Sort_j^{j\in J}
\overline{\pNet}\!,
\overline{\symb{SV}}\myrangle)
=\{\alpha' |\, \overline{\alpha}%\alpha_j^{j\in J_k}
\to\alpha'\in\set{\symb{SV}}\}
\end{array}
\]

>>>>

\section{General Notations and pNets previous results} 
\label{secn:preliminaries}

@@

\subsection{Notations}
\label{secn:notations}

We extensively use indexed structures
over some countable indexed sets, which are equivalent to mappings over
the countable set. 
Thus, 
$a_i^{i\in I}$
%, or equivalently  $(i\mapsto a_i)^{i\in I}$
denotes a family of elements $a_i$ indexed over the
set $I$. % Such a family
% is equivalent to the mapping $(i\mapsto a_i)^{i\in I}$.
% To specify the set over which the structure is indexed,
% indexed structures are always denoted with an exponent of the form $i\in I$
% (arithmetic only appears in the indexes if necessary).
This notation defines both $I$ the set over which the family is
indexed (called \emph{range}), and $a_i$ the elements of the family.
E.g., $a^{i\in\{3\}}$ is the mapping with a single entry $a$ at index
$3$ ; also abbreviated $(3\mapsto a)$.
When this is not
ambiguous, we shall use notations for sets, and typically write
``indexed set over I'', even though formally we should speak of multisets; and
write $x\in a_i^{i\in I}$ to mean $\exists i\in I.\, x=a_i$.  An empty
family is denoted $\emptyset$.

%\noteIn{states looks ok, but initial state is $s_0$ in pNet and not in archi : $s^0$.  Exponent 
%position was removed from pNet because we found it complex with indexed set notation but 
%could be ok.
%\Ludo
%
%I am using exponent notation for ``modifiers'', reserving\mdash
%as much as possible\mdash index notation for enumeration.  I
%think this can be solved, where necessary, by using parentheses,
%\eg $(s^0_i)^{i \in I}$
%\Simon
%}

%****************************************************************
We assume the existence of a term algebra $\Talg$,
where $\signature$ is the signature of the data and action constructors,
and $\variables$ a set of \emph{variables}. Within $\Talg$, we distinguish a set of
\emph{data expressions} $\exprs{\variables}$; $e_i$ ranges over boolean expressions, and a set of \emph{Boolean
expressions} 
$\boolexprs{\variables}\subseteq\exprs{\variables}$, $g$ (guards) ranges over boolean expressions.
On top of $\exprs{\variables}$ we build the \emph{action algebra}
$\actions{\variables}$, with $\actions{\variables}\subseteq\Talg$.
We  use
\[
\assigns{\variables} \bydef{=}
\Setdef{(x_i := e_i)^{i \in I}}{
  x_i^{i \in I}\subseteq \variables,
  e_i^{i \in I}\subseteq \exprs{\variables},
  I \text{ is a finite index set}
}
\]
to denote the set of \emph{variable assignments}.  $u$ ranges over sets of assignments.
To be able to reason about the data flow between actions, we
distinguish \emph{input variables} of the form $?x$ within terms; the function
$\vars(t)$ identifies the set of variables in a term
$t\in\Talg$, and $iv(t)$ returns its input variables.

We assume the existence of a universal data domain
given as a partially-ordered set $(\data, \order)$, potentially
encompassing several copies of any given data type with
different orders.  We assume that $(\data,
\order)$ comprises both the unordered set of Booleans $\sB =
\tuple{\{\true, \false\}, \emptyset}$ and the naturally ordered one
$\sB^< = \tuple{\{\true, \false\}, \{\false < \true\}}$, and
similarly for integer and real numbers; as well as the set of
intervals ordered by inclusion.
%In practice, and in the examples of
%this paper, all the variables are sorted, meaning that their values
%can only range over given subsets of $\data$, such as $\sB$ or
%$\sB^<$.
%
When speaking of an ordered sort, \eg $\sB^<$, we will assume that it
forms a meet-semilattice and denote by $\wedge$ the meet
operator.

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$.
Valuations extend canonically to expressions and denote $\val{}{}(e)$ the value of $e$ under the valuation $\val{}{}$.
\noteLH{its the following really necessary}
\addSB{Since we consider variables to be sorted, valuations
  applied to expressions are interpreted as partial functions
  $\exprs{V} \rightharpoonup \data$.}
For a valuation $\val{}{}$ and an assignment $(x_i := e_i)^{i \in I}$, we define:
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e$ is  \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} $ is
 monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We  denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%
%****************************************************************

@@

%****************************************************************
%****************************************************************
%****************************************************************

\section{The theory of architectures with data}
\label{secn:archs}

@@

\bibliographystyle{abbrv}
\bibliography{biblio,biblioFromPnets}

@@

\begin{definition}[Sorts, Holes, Leaves of pNets]
  \begin{itemize}
  \item  \noteSB{I think this can be lightened up a bit, since the definitions are all 
  straightforward, whereas the notation is complex.  Maybe reformulate in words and move 
  the formal definitions in the appendix?}
 The sort of a pNet is its signature, i.e. the set of actions it can
perform. \noteSB{Is there any chance that, by writing assignments explicitly ($x := e$), 
we could eliminate the need to distinguish the input variables at all? That would 
simplify things.}\hl{In the definition of sorts, we do not need to distinguish
input variables} (that specify the dataflow within LTSs), so for
computing LTS sorts, we use a substitution operator\footnote{$\subst{y_k\gets x_k}^{k\in 
K}$ is the parallel substitution 
operation.} to remove the
\emph{input marker} of variables\footnote{$\overline{S}$ denotes a set of $S$.}. Formally:
\[
\begin{array}{l}
\Sortop(\mylangle S,s_0, \to\myrangle) = \{\alpha\subst{x \gets ?x| 
x\in\symb{iv}(\alpha)}|s \xrightarrow{\langle \alpha,~g,~(x_j\!:= {e}_j)^{j\in
    J}\rangle} s'\in \to \} \\
\Sortop(\mylangle \overline{\pNet}\!, %\pNet_i^{i\in I}, \Sort_j^{j\in J}
\overline{\pNet}\!,
\overline{\symb{SV}}\myrangle)
=\{\alpha' |\, \overline{\alpha}%\alpha_j^{j\in J_k}
\to\alpha'\in\set{\symb{SV}}\}
\end{array}
\]
